### 2025-04-26 paper/ecai2025/sects/experiments.tex
minors

In this section, we test \fname{} with PPO and SHAC for a set of multi-agent scenarios and architectures. An ablation study when replacing only the transition function (SHAC+,meaning that the reward function is assumed differentiable) is also conducted in \Cref{apx:ablation}.

\paragraph{Scenarios}
\begin{figure*}
    \centering
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion.png}
        \caption{Dispersion}
        \label{fig:dispersion}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/transport.png}
        \caption{Transport}
        \label{fig:transport}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/discovery.png}
        \caption{Discovery}
        \label{fig:discovery}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/sampling.png}
        \caption{Sampling}
        \label{fig:sampling}
    \end{subfigure}
    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}\vspace{0.5cm}
\end{figure*}
We evaluate our algorithms on several multi-agent scenarios from VMAS~\cite{DBLP:conf/dars/BettiniKBP22}, which we selected over alternatives like PettingZoo~\cite{DBLP:conf/icml/PettingZoo}, SMAC~\cite{DBLP:conf/icml/ZhangZLZ20}, MA-MuJoCo~\cite{DBLP:conf/aaai/GuptaK0G20}, and others due to its key advantages: differentiable physics (enabling gradient-based optimization), multi-agent design, vectorized implementation (for efficient parallel training), realistic physics-based simulation, and open-source accessibility.
For more details about the choice of environments, see \Cref{apx:scenarios}.
We selected four distinct scenarios: Dispersion, Discovery, Transport, and Sampling, specifically for their cooperative nature and varying levels of required coordination (see \Cref{fig:scenarios} and \Cref{apx:scenarios} for details). These scenarios can be categorized along two dimensions:
\begin{compactitem}
    \item \textbf{Reward differentiability:} Dispersion and Discovery feature non-differentiable reward functions (making SHAC inapplicable), while Transport and Sampling have differentiable rewards (allowing for SHAC implementation).
    \item \textbf{Observation completeness:} Transport and Dispersion provide agents with complete observation spaces relative to the environment state. In contrast, Discovery and Sampling have incomplete observation spaces, which presents an additional challenge for \fname{} compared to SHAC, as \fname{} lacks access to the complete environmental state.
\end{compactitem}

@@

\subsection{Results}
In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of achieved rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.

@@

\paragraph{Transport}
The third row of \Cref{fig:experiments} presents a differentiable scenario, transport. In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance.\ \fname{} converges faster, but SHAC is more consistent. In the second and third columns (3 and 5 agents), PPO fails as in the single-agent case, while SHAC and \fname{} both complete the task around 10,000 episodes. With more agents, convergence is faster because the task becomes easier (as more agents can push the package). The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package. This is a sign of generalization, as the agents learn to avoid collisions and push the package together.

>>>>

In this section, we evaluate \fname{} against both PPO and SHAC across various multi-agent scenarios and architectures. An ablation study replacing only the transition function (SHAC+, where the reward function remains differentiable) is provided in \Cref{apx:ablation}.
We select PPO as our baseline because, despite the emergence of more recent algorithms HAPPO \cite{Kuba2021TrustRM}, and SAC \cite{Haarnoja2018SoftAA}, it remains one of the few approaches that scales effectively to multi-agent settings without substantial modifications (e.g., MAPPO~\cite{DBLP:conf/nips/YuVVGWBW22}). This makes PPO a representative benchmark for state-of-the-art performance in non-differentiable environments.

\paragraph{Scenarios}
We evaluate our algorithms on several multi-agent scenarios from VMAS~\cite{DBLP:conf/dars/BettiniKBP22}, which we selected over alternatives like PettingZoo~\cite{DBLP:conf/icml/PettingZoo}, SMAC~\cite{DBLP:conf/icml/ZhangZLZ20}, MA-MuJoCo~\cite{DBLP:conf/aaai/GuptaK0G20}, and others due to its key advantages: differentiable physics (enabling gradient-based optimization), multi-agent design, vectorized implementation (for efficient parallel training), realistic physics-based simulation, and open-source accessibility.
For more details about the choice of environments, see \Cref{apx:scenarios}.
We selected four distinct scenarios: Dispersion, Discovery, Transport, and Sampling, specifically for their cooperative nature and varying levels of required coordination (see \Cref{fig:scenarios} and \Cref{apx:scenarios} for details). These scenarios can be categorized along two dimensions:
\begin{compactitem}
    \item \textbf{Reward differentiability:} Dispersion and Discovery feature non-differentiable reward functions (making SHAC inapplicable), while Transport and Sampling have differentiable rewards (allowing for SHAC implementation).
    \item \textbf{Observation completeness:} Transport and Dispersion provide agents with complete observation spaces relative to the environment state. In contrast, Discovery and Sampling have incomplete observation spaces, which presents an additional challenge for \fname{} compared to SHAC, as \fname{} lacks access to the complete environmental state.
\end{compactitem}

@@

\subsection{Results}
\begin{figure}
    \centering
    \includegraphics[width=0.32\columnwidth]{figs/step-1.png}
    \includegraphics[width=0.32\columnwidth]{figs/step-2.png}
    \includegraphics[width=0.32\columnwidth]{figs/step-3.png}
    \caption{Snapshots of the result of the Transport scenario. The agents are represented as colored circles, the package as a red square, and the goal as a green circle. The agents are trained to push the package to the goal.}\label{fig:transport-trained}
    \vspace{0.5cm}
\end{figure}
In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of achieved rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.

@@

\paragraph{Transport}
The third row of \Cref{fig:experiments} presents a differentiable scenario, transport. In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance.\ \fname{} converges faster, but SHAC is more consistent. In the second and third columns (3 and 5 agents), PPO fails as in the single-agent case, while SHAC and \fname{} both complete the task around 10,000 episodes. With more agents, convergence is faster because the task becomes easier (as more agents can push the package). The emergent of cooperation (see \Cref{fig:transport-trained}) can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package. This is a sign of generalization, as the agents learn to avoid collisions and push the package together.
